Dependent Object Types

### 2012-09-22 doc/fool/article.tex
type safety to type-safety and typecheck to type-check

The expansion relation $\expand$ is needed to typecheck the complete
set of declarations carried by a concrete type that is used in a
\textbf{new}-expression. Expansion is also used by the membership
rules and in subtyping refinements on the right (see
\figref{dot-sub}).

@@

$a$ enforces that its first argument be a $\mi{pets.Poodle_c}$ and its second a $\mi{pets.Dalmatian_c}$. The following does not typecheck for this reason:
\begin{align*}
\cast \Top {\app {a.choose(d)} p}
\end{align*}

@@

More generally, these counterexamples illustrate that preservation
doesn't hold because a term that typechecks can step to a term that
does not typecheck. However, these counterexamples to preservation are
not counterexamples to type-safety: i.e. these programs don't get
stuck -- they eventually step to a value.

@@

Here is the simplest such example:
\begin{align*}
&\mlnew z {\Top \mlrefine z {\\
&\gap \Ldecl {L} \Bot {z.L}\\
&}\ldefs{}}{}
\end{align*}
The type $z.L$ is $\nswf$ but not $\nswfe$. Indeed, there is no finite
derivation of an expansion for $z.L$, because by the
$\textsc{tsel-$\expand$}$ rule, in order to expand $z.L$, we need to
expand its upper bound, which is $z.L$! However, not that the object
creation expression for $z$ would not typecheck because subtyping is
regular wrt $\nswfe$, and so the type of a constructor in an object
creation expression must have $\nswfe$ type members, since we check
that the lower bound is a subtype of the upper bound for each type
member.

@@

\section{Type Safety via Logical Relations}\label{dot-type-safety}

We believe that the DOT calculus is type-safe, and are developing a
proof of type safety using step-indexed logical
relations~\cite{ahmed04,ahmed06,step_indexed_obj}. In this section, we
briefly summarize the main theorem of type safety, and the strategy
based on step-indexed logical relations that we are using to prove
it. All our development, including models in Coq~\cite{coq} and PLT Redex~\cite{plt_redex}, is
available from \url{http://dot.namin.net}.

\subsection{Type Safety}
Type safety states that a well-typed program doesn't get stuck. More
formally: If $\emptyset \ts t \typ T$ and $\reductionl t \emptyset
{t'} {s'} *$ then either $t'$ is a value or $\exists t'',
s''. \reduction {t'} {s'} {t''} {s''}$. Note that this is stronger
than the standard theorem of progress, which states that a well-type
term can take \emph{a} step or is a value.

\subsection{Strategy based on Logical Relations}
Our strategy follows the standard technique of proving type safety
using logical relations. We define a logical relation $\Gamma \ds t
: T$, such that $\Gamma \ts t : T$ implies $\Gamma \ds t : T$ implies
type safety.

@@

Then, to prove type safety, all that needs to be proved is the
fundamental theorem -- or completeness -- of the logical relation:
$\Gamma \ts t : T$ to $\Gamma \ds t : T$. Type safety is a
straightforward corollary of this theorem, since $\Gamma \ds t : T$
implies by definition that $t$ cannot be stuck. The proof of the
fundamental theorem is on induction on the derivation of $\Gamma \ts t
: T$. The logical relation approach enables us to have strong enough
induction hypotheses to carry the proof through, without requiring us
to strictly relate intermediate terms by types like preservation.

@@

In the calculus we made the deliberate choice not to model any form of
inheritance. This is, first and foremost, to keep the calculus simple.
Secondly, there are many different approaches to inheritance and
mixin composition, so that it looks advantageous not to tie the basic
calculus to a specific one. Finally, it seems that the modelization of
inheritance lends itself to a different approach than the basic
calculus. For the latter, we need to prove type safety of the calculus.
One might try to do this also for a calculus with inheritance, but our
experience suggests that this complicates the proofs considerably.  An
alternative approach that might work better is to model inheritance
as a form of code-reuse. Starting with an enriched type system with
inheritance, and a translation to the basic calculus, one needs to
show type safety wrt the translation. This might be easier than
to prove type safety wrt reduction.

@@

The idea is to start with a function from $S \tfun S$ and cast it
successively to $S \tfun T$ then $S \tfun U$. To typecheck the
expression initially, we need to check $S \sub T$ and $T \sub
U$. After some reduction steps, the first few casts vanish, and the
reduced expression casts directly from $S \tfun S$ to $S \tfun U$, so
we need to check $S \sub U$.

@@

Now, $f$ was a subtype of $\Top \tfun \Top$, but $\app f {(\abs x \Top
  x)}$ was stuck (and, rightfully, didn't typecheck). But we could use
narrowing to create a counterexample to type safety: $\app {(\abs g
  {\Top \tfun \Top} {\app g (\abs x \Top x)})} f$.

@@

Here is a summary of one attempt, which is further detailed
below. Because many of the counterexamples to preservation are re-
lated to narrowing, we tried to make widening an explicit operation
and change rules with implicit relaxations (MSEL and NEW) to be
strict. From a typing perspective, the change was straightforward, but
reduction became more complicated and dependent on typing because the
type information in widenings needed to be propagated correctly.  We
added path equality provisions in the subtyping rules, in the same
spirit as the Tribe calculus~\cite{tribe}.  Unfortunately, these two
patches interacted badly, and we were left with a disturbing
counterexample to type safety.

@@

Notice that $d$ has type $\Bot$ if you ignore the cast on $b$. This
example doesn't typecheck initially because $\textsc{path-$\sub$}$
only applies when objects are in the store, so the application is not
well-typed. But if we start preservation in a store which has $a$,
$b$, $c$ and $d$ then the application type-checks, because, through
$\textsc{path-$\sub$}$, we can find that the type of $d$ is a subtype
of $\Bot$. Now, of course, when we get to $d.\mathit{foo}$, reduction
fails.

@@

Proving the DOT calculus type-safe has been an interesting
adventure. We have shown that DOT does not satisfy preservation (also
known as subject-reduction), because a well-typed term can step to an
intermediate term that is not well-typed. In any case, the standard
theorems of preservation and progress are just one way to prove type
safety, which states that a well-typed term cannot be stuck -- a
weaker statement that does not require typechecking intermediate
steps. We are developing a proof of type safety using logical
relations.

>>>>

The expansion relation $\expand$ is needed to type-check the complete
set of declarations carried by a concrete type that is used in a
\textbf{new}-expression. Expansion is also used by the membership
rules and in subtyping refinements on the right (see
\figref{dot-sub}).

@@

$a$ enforces that its first argument be a $\mi{pets.Poodle_c}$ and its second a $\mi{pets.Dalmatian_c}$. The following does not type-check for this reason:
\begin{align*}
\cast \Top {\app {a.choose(d)} p}
\end{align*}

@@

More generally, these counterexamples illustrate that preservation
doesn't hold because a term that type-checks can step to a term that
does not type-check. However, these counterexamples to preservation are
not counterexamples to type-safety: i.e. these programs don't get
stuck -- they eventually step to a value.

@@

Here is the simplest such example:
\begin{align*}
&\mlnew z {\Top \mlrefine z {\\
&\gap \Ldecl {L} \Bot {z.L}\\
&}\ldefs{}}{}
\end{align*}
The type $z.L$ is $\nswf$ but not $\nswfe$. Indeed, there is no finite
derivation of an expansion for $z.L$, because by the
$\textsc{tsel-$\expand$}$ rule, in order to expand $z.L$, we need to
expand its upper bound, which is $z.L$! However, not that the object
creation expression for $z$ would not type-check because subtyping is
regular wrt $\nswfe$, and so the type of a constructor in an object
creation expression must have $\nswfe$ type members, since we check
that the lower bound is a subtype of the upper bound for each type
member.

@@

\section{Type-Safety via Logical Relations}\label{dot-type-safety}

We believe that the DOT calculus is type-safe, and are developing a
proof of type-safety using step-indexed logical
relations~\cite{ahmed04,ahmed06,step_indexed_obj}. In this section, we
briefly summarize the main theorem of type-safety, and the strategy
based on step-indexed logical relations that we are using to prove
it. All our development, including models in Coq~\cite{coq} and PLT Redex~\cite{plt_redex}, is
available from \url{http://dot.namin.net}.

\subsection{Type-Safety}
Type-safety states that a well-typed program doesn't get stuck. More
formally: If $\emptyset \ts t \typ T$ and $\reductionl t \emptyset
{t'} {s'} *$ then either $t'$ is a value or $\exists t'',
s''. \reduction {t'} {s'} {t''} {s''}$. Note that this is stronger
than the standard theorem of progress, which states that a well-type
term can take \emph{a} step or is a value.

\subsection{Strategy based on Logical Relations}
Our strategy follows the standard technique of proving type-safety
using logical relations. We define a logical relation $\Gamma \ds t
: T$, such that $\Gamma \ts t : T$ implies $\Gamma \ds t : T$ implies
type-safety.

@@

Then, to prove type-safety, all that needs to be proved is the
fundamental theorem -- or completeness -- of the logical relation:
$\Gamma \ts t : T$ to $\Gamma \ds t : T$. Type-safety is a
straightforward corollary of this theorem, since $\Gamma \ds t : T$
implies by definition that $t$ cannot be stuck. The proof of the
fundamental theorem is on induction on the derivation of $\Gamma \ts t
: T$. The logical relation approach enables us to have strong enough
induction hypotheses to carry the proof through, without requiring us
to strictly relate intermediate terms by types like preservation.

@@

In the calculus we made the deliberate choice not to model any form of
inheritance. This is, first and foremost, to keep the calculus simple.
Secondly, there are many different approaches to inheritance and
mixin composition, so that it looks advantageous not to tie the basic
calculus to a specific one. Finally, it seems that the modelization of
inheritance lends itself to a different approach than the basic
calculus. For the latter, we need to prove type-safety of the calculus.
One might try to do this also for a calculus with inheritance, but our
experience suggests that this complicates the proofs considerably.  An
alternative approach that might work better is to model inheritance
as a form of code-reuse. Starting with an enriched type system with
inheritance, and a translation to the basic calculus, one needs to
show type-safety wrt the translation. This might be easier than
to prove type-safety wrt reduction.

@@

The idea is to start with a function from $S \tfun S$ and cast it
successively to $S \tfun T$ then $S \tfun U$. To type-check the
expression initially, we need to check $S \sub T$ and $T \sub
U$. After some reduction steps, the first few casts vanish, and the
reduced expression casts directly from $S \tfun S$ to $S \tfun U$, so
we need to check $S \sub U$.

@@

Now, $f$ was a subtype of $\Top \tfun \Top$, but $\app f {(\abs x \Top
  x)}$ was stuck (and, rightfully, didn't type-check). But we could use
narrowing to create a counterexample to type-safety: $\app {(\abs g
  {\Top \tfun \Top} {\app g (\abs x \Top x)})} f$.

@@

Here is a summary of one attempt, which is further detailed
below. Because many of the counterexamples to preservation are re-
lated to narrowing, we tried to make widening an explicit operation
and change rules with implicit relaxations (MSEL and NEW) to be
strict. From a typing perspective, the change was straightforward, but
reduction became more complicated and dependent on typing because the
type information in widenings needed to be propagated correctly.  We
added path equality provisions in the subtyping rules, in the same
spirit as the Tribe calculus~\cite{tribe}.  Unfortunately, these two
patches interacted badly, and we were left with a disturbing
counterexample to type-safety.

@@

Notice that $d$ has type $\Bot$ if you ignore the cast on $b$. This
example doesn't type-check initially because $\textsc{path-$\sub$}$
only applies when objects are in the store, so the application is not
well-typed. But if we start preservation in a store which has $a$,
$b$, $c$ and $d$ then the application type-checks, because, through
$\textsc{path-$\sub$}$, we can find that the type of $d$ is a subtype
of $\Bot$. Now, of course, when we get to $d.\mathit{foo}$, reduction
fails.

@@

Proving the DOT calculus type-safe has been an interesting
adventure. We have shown that DOT does not satisfy preservation (also
known as subject-reduction), because a well-typed term can step to an
intermediate term that is not well-typed. In any case, the standard
theorems of preservation and progress are just one way to prove type
safety, which states that a well-typed term cannot be stuck -- a
weaker statement that does not require type-checking intermediate
steps. We are developing a proof of type-safety using logical
relations.
